The Power of Unbiased Recursive Partitioning:\\ A Unifying View of CTree, MOB, and GUIDE

### 2019-01-16 pkg/partykit/inst/guideliketest/draft/draft.Rnw
further revisions, added references and added function to prepare data set for 3-way facotrial analysis
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2405 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 94e669fcfa7e6c18b775c493b98b53d55df26910

%% packages
\usepackage{amstext,amsfonts,amsmath,amssymb,bm,thumbpdf,lmodern,hyperref}
\usepackage[all]{hypcap}

@@

## HCL palette
pal <- hcl(c(140, 80, 260, 70, 10), 300, 60)
names(pal) <- c("CTree", "CTree_max", "MOB", "GUIDE", "GUIDE_full")

@@

Therefore, three different approaches are considered here and embedded into a common 
modeling framework with special emphasis to linear model trees: classical categorical 
association tests (GUIDE), conditional inference (CTree), parameter instability 
tests (MOB). It is assessed how different building blocks affect the power of the 
tree algorithms to select the appropriate covariates for splitting: residuals vs. full 
model scores, binarization of residuals/scores at zero, binning of covariates, 
conditional vs.\ unconditional approximations of the null distribution.
}
\Keywords{classification and regression trees, independence tests, recursive partitioning}

@@

\section{Introduction}\label{sec:introduction}
In many situations fitting one global model to a data set can be very challenging, 
especially if this data contains lots of different features with high/extreme variations 
or complex interactions. Therefore, separating the data into more homogeneous subgroups 
based on (a set of) covariates first can simplify the task and fitting a local model to 
each of the resulting subgroups often leads to better results. This separation of the 
data can be done by applying a tree algorithm. Many different algorithms have been developed
in this research field. While they all follow the general idea of splitting the data such 
that some objective function is minimized/maximized, they differ in their 
specific/particular approaches to selecting a split variable and the corresponding split point. 
Some of the first tree algorithms (e.g., AID \citealp{Morgan+Sonquist:1963}; 
CART \citealp{Breiman+Friedman+Stone:1984}) rely on exhaustive search procedures to find both, 
the best split point and split variable in one step by directly comparing all possible split
points in all possible split variables. However, it has been shown that this is not only
computationally expensive but also biased towards split variables with many possible split
points \citep{Doyle:1973, Kim+Loh:2001}. 
Therefore, selecting a split variable in a first step and then searching for the 
best split point only within this variable in a separated second step is a more 
promising strategy as applied for example by the algorithms QUEST \citep{Loh+Shih:1997},
CHAID \citep{Kass:1980}, GUIDE \citep{Loh:2002}, CTree \citep{Hothorn+Hornik+Zeileis:2006} 
and MOB \citep{Zeileis+Hothorn+Hornik:2008}.
For the first step of selecting a split variable they all follow the idea of 
choosing the covariate which shows the highest association to/influence on the response variable 
based on a statistical test. While QUEST, CHAID and GUIDE employ statistical significance
tests for contingency tables, CTree applies permutation tests in a conditional inference
framework and MOB uses fluctuation tests based on asymptotic theory.

In this paper the focus will be put on this first step, i.e. the task of selecting 
the best split variable. In particular, the approach of the GUIDE algorithm is
compared to the one of the CTree algorithm and the MOB algorithm. For this purpose, a 
unifying framework for unbiased model-based tree algorithms is presented such that each 
of the three algorithms can be obtained by a specific combination of the available building 
blocks. In that way, the effects of each of these characteristic features of the GUIDE,
CTree and MOB algorithm can be investigated separately by varying the switches in
the tool box provided by this general framework. 
Depending on the setup different effects can be observed for the investigated characteristics.
However, none of the results contradicts the conclusion that in general using all available 
model scores is at least as efficient but mostly clearly advantageous over considering only 
residuals. Moreover, the effects of binarizing residuals or scores at $0$ or categorizing
the values of possible split variables are investigated and the impact of applying a conditional
or an unconditional approximation of the null distribution is analyzed.

In the first part of Section~\ref{sec:urp} a basic scheme on how to build a tree is given. 
Then, after explaining the investigated testing strategies more in detail in 
Sections~\ref{sec:ctree}, \ref{sec:mob} and \ref{sec:guide}, the unifying framework for 
model-based unbiased recursive partitioning is presented in Section~\ref{sec:unify}.
The setting for the simulation study is introduced in Section~\ref{sec:simulation}
and the results are discussed in Section~\ref{sec:results}.

@@

\subsection{Generic Algorithm}
The basic idea of building a regression tree model is to partition the data into smaller and 
more homogeneous subgroups based on (a set of) covariates. Various different tree algorithms 
have been developed, all following the same general structure starting at the root of the tree
which is simply the first node containing all available data:
\begin{enumerate}
\item For the whole underlying data set in the current node some form of discrepancy 
measure is calculated. 
%This can for example be the sum of deviations from a typical or average value of the 
%data set or it can be based on a fitted model.
\item Among all available covariates one is selected as split variable and a value 
within the range/measurable space of this variable is chosen as split point such that 
after splitting the data into subgroups the discrepancy measure is minimal within the 
resulting nodes.
\item Steps 1 and 2 are repeated within each new node until either a stopping criterion
is reached or the data in the node consists of only one or all equal observations. 
\end{enumerate}

The employed discrepancy measure can for example be the sum of deviations from a typical 
or average value, but it can also be constructed based on a statistical model. For instance,
different types of residuals or their signs can be applied (see e.g., \cite{Loh:2002}) as well as
rank sums or logrank scores as in \cite{Hothorn+Hornik+Wiel:2006}.
(In the simulation study of this paper trees with linear models fitted in each node are 
investigated with the residual sum of squares (RSS) employed as discrepancy measure.)

As explained in Section~\ref{sec:introduction} the considered tree algorithms CTree, MOB, and 
GUIDE first select the split variable and then the split point in a separate step. 
For the first step of selecting a split variable they all apply statistical tests
following the same general strategy: 
Based on the discrepancy measure an $n \times k$-matrix $D$ is obtained giving information
on the goodness of fit of the model for a data set of $n \in \mathbb{N}$ observations in the 
current node and $k \in \mathbb{N}$.
Using this matrix, the algorithm-specific test statistic is calculated for each possible split 
variable $Z_j$, $j \in \{1,\ldots,m\}$ for $m \in \mathbb{N}$, evaluating the dependency
of $Z_j$ on $D$. As the variables are usually not all measured at the same scale the resulting
values of the test statistics can not be compared directly in an unbiased way. Therefore, it is
necessary to calculate the corresponding $p$-values in order to allow for an unbiased
variable selection on this standardized/unified scale. Then the variable corresponding to the smallest
$p$-value is selected as split variable as it has shown to be the variable with the strongest
influence on the goodness of fit of the model.

While this basic approach is the same for the GUIDE, the CTree and the MOB algorithm they differ 
in their strategies on how to calculate test statistics. In order to point out these specific 
characteristics in Section~\ref{sec:unify} the strategies of the three tree algorithms are first 
explained in detail in Sections~\ref{sec:ctree}.

@@

The CTree algorithm is based on the idea of providing non-parametric regression tree models in 
a conditional inference framework by applying permutation tests. 
To select a split variable it is tested whether there is any association between the transformed 
response $h(Y)$ and $g(Z)$ for a so-called influence function $g$ and each possible split variable $Z$.
For different types of response variables different functions $h$ which depend on $Y$ in a 
permutation-symmetric way and might also be multidimensional can be selected.
For a numeric response the identity function $h(Y) = Y$ is a common choice, while a categorical
response can be mapped to a unity vector by an indicator function $h(Y) = (0,\ldots,1,\ldots,0)^{\top}$.
Alternatively, the function $h$ can return a location and scale parameter of $Y$, 
e.g. $h(Y) = (Y, (Y-\bar{Y})^2)^{\top}$. 
If a parametric model is fit to the values of the response variable~$Y$ and regressor variable(s)~$X$
the function $h$ can be a model-based transformation returning the corresponding model scores by 
setting $h(Y) = s(Y,X,\hat{\theta})$
where $\hat{\theta}$ is the estimated model parameter and $s$ is the score function
$$
s(Y, X, \hat{\theta}) = \frac{\partial \psi(Y, X, \hat{\theta})}{\partial \theta}
$$
with $\psi$ being the objective function which is minimized to estimate the model parameter
$\theta$. Hence, for a data set of $n$ observations containing the response vector 
$\boldsymbol{y} = (y_1,\ldots,y_n)^\top$ and 
regressor(s) $\boldsymbol{x} = (x_1,\ldots,x_n)^\top$ and 
a $k$-dimensional parameter vector $\boldsymbol{\theta} = (\theta_1,\ldots,\theta_k)^\top$ the 
score function $s$ returns a $n \times k$-matrix.
In that way a score value indicating how well the model fits the data is given separately
for each tuple of observation and parameter element $(y_i, x_i, \theta_l)$ with $i \in \{1,\dots,n\}$
and $l \in \{1,\ldots,k\}$. 
To estimate $\boldsymbol{\theta}$ many different M-type estimators can be applied including 
maximum likelihood or ordinary least squares. Therefore, these objective functions
are also possible options to obtain a model-based discrepancy measure in the tree algorithm.

@@

If both variables $Y$ and $Z$ are numeric the applied independence test corresponds to a Pearson 
correlation test. For one of the variables being categorical and the other numeric a 1-way ANOVA
is employed while for two categorical variables a $\chi^2$-test is performed. Hence, the CTree algorithm
provides a very general framework where the choice of the type of variables and the corresponding 
functions $g$ and $h$ determine the type of the applied statistical test.

@@

\subsection{MOB}\label{sec:mob}
While CTree offers to apply model-based discrepancy measures as one possible choice, the MOB algorithm 
was developed with the aim of embedding parametric models into the framework of regression trees
and is hence purely intended for a model-based approach. The testing strategy for selecting a split
variable is based on a fitted model, i.e. on an estimated parameter (vector) $\hat{\theta}$ for each
node of the tree, and on the application of the central limit theorem. However, the fitted parametric 
models can be of a very basic and simple form such as an average value. The algorithm only requires the 
corresponding model scores $s(Y,X,\hat{\theta})$ in order to test for any instability in the model 
parameters over each particular ordering induced by each of the possible partitioning variables $Z$.
If a variable $Z$ has no specific influence on the parameters the scores would always fluctuate randomly
around their mean 0. Therefore, these deviations are investigated by measuring them using an 
empirical fluctuation process. Applying a scalar function which captures the fluctuation in this process 
yields the test statistic of this so called generalized M-fluctuation test. To obtain the distribution of
the resulting test statistic asymptotic theory, in particular the central limit theorem is applied.
In this parametric unconditional framework $p$-values can be calculated and the covariate
with the smallest $p$-value is selected as split variable. 
For a categorical variable all possible splits are evaluated to find the best one while for $Z$ being 
a numeric variable a maximally selected split point search is employed. Hence, the number 
of possible breakpoints is not fixed unlike in the CTree algorithm which can make this operation 
computationally expensive.

\subsection{GUIDE}\label{sec:guide}
Similar to the MOB algorithm, parametric models are fit to the nodes of a tree in the GUIDE algorithm.
For the statistical test applied to select a split variable in each node some type of 
residuals from the models is required such as for example 
$r(Y,X,\hat{\theta}) = Y - X^{\top}\hat{\theta}$. 
Before applying a statistical test two additional steps are performed:
The residuals are binarized, i.e. depending on whether the sign of a residual is 
positive or not it is replaced by $1$ or $2$  % FIX ME: +1 and -1 in the implementation? 
and for each possible split variable the observed values are categorized into 4 groups 
separated by the corresponding 1st, 2nd and 3rd quartile.
Then, a $\chi^2$-test of independence is performed for the binarized residuals and
each categorized split variable.
After choosing the split variable showing the highest dependency by yielding the lowest $p$-value,
the split point minimizing the discrepancy measure is selected. The possible number of breakpoints
is determined by the categorization of the possible split variables such that the quartiles are
the only three possible split points.

\bigskip

\textit{Note:}
While GUIDE originally constructs large trees and applies post-pruning with the cross-validation 
method of CART, CTree and MOB employ pre-pruning by setting a significance level for the 
independence tests such that no split is performed if the lowest $p$-value is not smaller than 
the significance level. As explained in Section~\ref{sec:evaluation}, within the simulation study 
in Section~\ref{sec:simulation} all three algorithms are applied using pre-pruning in order to put 
the focus on the power of the algorithms in terms of selecting the best split variable. 
However, the impact of the choice of pruning strategy is further discussed in Section~\ref{sec:discussion}.

\section{Unifying Framework}\label{sec:unify}
Since all three algorithms follow the same basic structure their approaches can be presented together
in a unifying framework for a model-based tree algorithm from which each specific algorithm can be 
obtained by switching on or off certain features. By applying such a general framework each of the 
characteristic features of the three algorithms and its impact can be investigated in detail. 
These switches are explained in this Section and their setting for the algorithms
CTree, MOB, and GUIDE are listed in Table~\ref{tab:switches}.

\subsection{Switches}
As explained before, the selection of a split variable is based on a statistical test 
investigating the dependency of the response $Y$ on each possible split variables $Z$.
The three following switches all refer to the way how the given data is prepared/transformed
before employing such a statistical test.
\begin{itemize}
\item  Transformations of Y:\\
A model-based transformation of $Y$ is applied to obtain a discrepancy measure giving 
information about the goodness of fit of the model. This can be residuals $r(Y,X,\hat{\theta})$
as in the GUIDE algorithm or scores $s(Y,X,\hat{\theta})$ as in the MOB algorithm or, 
more generally, a function $h(Y,X)$ as in the CTree algorithm including both of the foregoing cases.
Since the first column of the score matrix $S = s(Y,X,\hat{\theta})$ is the vector of residuals the 
GUIDE algorithm selects only this first column while the MOB algorithm uses all available score values,
hence the full model score matrix $S$.
%providing information on the goodness of fit for each observation and model parameter.
(By including all available scores the goodness of fit of the model is investigated regarding 
each parameter rather than just regarding the mean/location parameter. In that way changes in 
parameters other than the mean/location parameter should be detected more easily.)

\item Binarization of scores:\\
Rather than applying the independence test directly on the selected (columns of) scores a
binarized version of them can be used as well. In particular, the score values are binarized
at $0$ such that only their signs are considered.

@@

\textit{Note:}
Apart from these three switches, the algorithms differ in their approximation of the underlying
null distribution of the test statistic. While the permutation tests in the CTree algorithm
work in a conditional framework the MOB and GUIDE algorithms both apply an unconditional approach.

\begin{table}[t!]
\centering
\caption[Table caption text]{CTree, MOB, and GUIDE with the corresponding setting of
the switches in the unifying framework and the approximation type of the underlying null distribution.}
\label{tab:switches}
\begin{tabular}{ l c c c c }
\hline 
& Scores            & Binarization  & Categorization  & Null distribution \\ 
\hline
CTree & full model scores & -           & -             & conditional\\
MOB   & full model scores & -           & -             & unconditional\\
GUIDE & residuals         & \checkmark  & \checkmark    & unconditional\\
\hline
\end{tabular}
\end{table}

>>>>

%% packages
\usepackage{amstext,amsfonts,amsmath,amssymb,bm,thumbpdf,lmodern,hyperref}
\usepackage[all]{hypcap}
\usepackage{enumitem}

@@

## HCL palette
pal <- hcl(c(140, 80, 260, 70, 10), 300, 60)
names(pal) <- c("CTree", "CTree_max", "MOB", "GUIDE", "GUIDE_scores")

@@

Therefore, three different approaches are considered here and embedded into a common 
modeling framework with special emphasis to linear model trees: classical categorical 
association tests (GUIDE), conditional inference (CTree), parameter instability 
tests (MOB). It is assessed how different building blocks affect the power of the 
tree algorithms to select the appropriate covariates for splitting: residuals vs.\ full 
model scores, binarization of residuals/scores at zero, binning of covariates, 
conditional vs.\ unconditional approximations of the null distribution.
}
\Keywords{classification and regression trees, independence tests, recursive partitioning}

@@

\section{Introduction}\label{sec:introduction}
In many situations fitting one global model to a data set can be very challenging, 
especially if this data contains lots of different features with high/extreme variations 
or complex interactions. Therefore, separating the data into more homogeneous subgroups 
based on (a set of) covariates first can simplify the task and fitting a local model to 
each of the resulting subgroups often leads to better results. This separation of the 
data can be done by applying a tree algorithm. Many different algorithms have been developed
in this research field. While they all follow the general idea of splitting the data such 
that some objective function is optimized, they differ in their 
specific approaches to selecting a split variable and the corresponding split point. 
Some of the first tree algorithms (e.g., AID \citealp{Morgan+Sonquist:1963}; 
CART \citealp{Breiman+Friedman+Stone:1984}) rely on exhaustive search procedures to find both, 
the best split point and split variable in one step by directly comparing all possible split
points in all possible split variables. However, it has been shown that this is not only
computationally expensive but also biased towards split variables with many possible split
points \citep{Doyle:1973, Kim+Loh:2001}. 
Therefore, selecting a split variable in a first step and then searching for the 
best split point only within this variable in a separated second step is a more 
promising strategy as applied for example by the algorithms QUEST \citep{Loh+Shih:1997},
CHAID \citep{Kass:1980}, GUIDE \citep{Loh:2002}, CTree \citep{Hothorn+Hornik+Zeileis:2006} 
and MOB \citep{Zeileis+Hothorn+Hornik:2008}.
For the first step of selecting a split variable they all share the same basic concept of 
choosing the covariate which shows the highest association to the response variable 
based on a statistical test. While QUEST, CHAID and GUIDE employ statistical significance
tests for contingency tables, CTree applies permutation tests in a conditional inference
framework and MOB uses fluctuation tests based on asymptotic theory. All these approaches
have shown to work well for various situations, however, the specific advantages and
disadvantages of each of the testing strategies have not yet been investigated and compared
in detail.

Therefore, in this paper the focus will be put on that first step of tree algorithms,
i.e. the task of selecting the best split variable. In particular, the approach of the 
GUIDE algorithm is compared to the one of the CTree algorithm and the MOB algorithm. 
For this purpose, a unifying framework for testing strategies in unbiased model-based tree 
algorithms is presented such that each of the three strategies can be obtained by a specific 
combination of the available building blocks. In that way, the effects of each of these 
features can be investigated separately by varying the building blocks in the tool box 
provided by this general framework.

Depending on the setup different effects can be observed for the investigated building blocks.
However, none of the results contradicts the conclusion that in general using all available 
model scores is at least as efficient but mostly clearly advantageous over considering only 
residuals. Moreover, the effects of binarizing residuals or scores at $0$ or categorizing
the values of possible split variables are investigated.
Moreover, the impact of applying either a pre- or postpruning strategy is analyzed.

In the first part of Section~\ref{sec:urp} a basic scheme on how to build a tree is given. 
Then, after explaining the investigated algorithms CTree, MOB, and GUIDE more in detail in 
Sections~\ref{sec:ctree}, \ref{sec:mob} and \ref{sec:guide}, the unifying framework for 
testing strategies in model-based unbiased recursive partitioning is presented in 
Section~\ref{sec:unify}.
The setting for the simulation study is introduced in Section~\ref{sec:simulation}
and the results are discussed in Section~\ref{sec:results}.

@@

\subsection{Generic Algorithm}
The basic idea of building a regression tree model is to partition the data into smaller and 
more homogeneous subgroups based on a set of covariates. Various different tree algorithms 
have been developed, all following the same general structure starting at the root of the tree
which is simply the first node containing all available data:
\begin{enumerate}
\item For the whole underlying data set in the current node some form of discrepancy 
measure is calculated. 
\item Among all available covariates one is selected as split variable and a value 
within the range/measurable space of this variable is chosen as split point such that 
after splitting the data into subgroups the discrepancy measure is minimal within the 
resulting nodes.
\item Steps 1 and 2 are repeated within each new node until either a stopping criterion
is reached or the data in the node consists of only one or all equal observations. 
\end{enumerate}

The employed discrepancy measure can for example be the sum of deviations from a typical 
or average value, but it can also be constructed based on a parametric model. For instance,
different types of residuals or their signs can be applied \citep[see e.g.,][]{Loh:2002} as 
well as rank sums or logrank scores as in \cite{Hothorn+Hornik+Wiel:2006}.
%(In the simulation study of this paper trees with linear models fitted in each node are 
%investigated with the residual sum of squares (RSS) employed as discrepancy measure.)

As explained in Section~\ref{sec:introduction} the considered tree algorithms CTree, MOB, and 
GUIDE first select the split variable and then in a separate step the split point. 
For the first step of selecting a split variable they all apply statistical tests
following the same general strategy: 
\begin{enumerate}
\item Based on the discrepancy measure an $n \times k$-matrix $D$ is obtained giving information
on the goodness of fit of the model for a data set of $n \in \mathbb{N}$ observations in the 
current node and $k \in \mathbb{N}$.
\item Using this matrix, the algorithm-specific test statistic is evaluated for each possible 
split variable $Z_j$, $j \in \{1,\ldots,m\}$ for $m \in \mathbb{N}$, assessing the dependency
of $Z_j$ on $D$ and in that way the dependency of $Z_j$ on the response $Y$ as the discrepancy
measure (and therefore also $D$) depends on $Y$. 
\item For the resulting values of the test statistics the corresponding $p$-values are calculated
in order to allow for their comparison on a standardized/unified scale and in that way for
an unbiased variable selection.
%As the variables are usually not all measured at the same scale the resulting
%values of the test statistics can not be compared directly in an unbiased way. Therefore, it is
%necessary to calculate the corresponding $p$-values in order to allow for an unbiased
%variable selection on this standardized/unified scale. 
\item The variable corresponding to the smallest $p$-value is selected as split variable as 
it has shown to be the variable with the strongest influence on the goodness of fit of the model.
\end{enumerate}

While this basic approach is the same for the GUIDE, the CTree and the MOB algorithm they differ 
in their strategies on how to calculate test statistics. In order to point out these specific 
characteristics in Section~\ref{sec:unify} the strategies of the three tree algorithms are first 
explained in detail in Sections~\ref{sec:ctree}, \ref{sec:mob}, and \ref{sec:guide} .

@@

The CTree algorithm \citep{Hothorn+Hornik+Zeileis:2006} is based on the idea of providing 
non-parametric regression tree models in a conditional inference framework by applying 
permutation tests. 
To select a split variable it is tested whether there is any association between the transformed 
response $h(Y)$ and $g(Z)$ for a so-called influence function $g$ and each possible split 
variable $Z$.
For different types of response variables different functions $h$ which depend on $Y$ in a 
permutation-symmetric way and might also be multidimensional can be selected.
For a numeric response the identity function $h(Y) = Y$ is a common choice, while a categorical
response can be mapped to a unity vector by an indicator function 
$h(Y) = (0,\ldots,1,\ldots,0)^{\top}$.
Alternatively, the function $h$ can return a location and scale parameter of $Y$, 
e.g. $h(Y) = (Y, (Y-\bar{Y})^2)^{\top}$. 
If a parametric model is fit to the values of the response variable~$Y$ and regressor 
variable(s)~$X$ the function $h$ can be a model-based transformation returning the 
corresponding model scores by setting $h(Y) = s(Y,X,\hat{\beta})$
where $\hat{\beta}$ is the estimated model parameter and $s$ is the score function
$$
s(Y, X, \hat{\beta}) = \frac{\partial \psi(Y, X, \hat{\beta})}{\partial \beta}
$$
with the sum of $\psi$ over all observations being the objective function that is optimized 
to estimate the model parameter $\beta$. For this purpose many different M-type estimators can 
be applied including maximum likelihood or ordinary least squares. Therefore, these objective 
functions are also possible options to obtain a model-based discrepancy measure for the tree 
algorithm.
Applying the score function $s$ on a data set of $n$ observations containing the 
response vector $\boldsymbol{y} = (y_1,\ldots,y_n)^\top$ and 
regressor(s) $\boldsymbol{x} = (x_1,\ldots,x_n)^\top$ and 
on a $k$-dimensional estimated parameter vector 
$\hat{\boldsymbol{\beta}} = (\hat{\beta_1},\ldots,\hat{\beta_k})^\top$ 
yields a $n \times k$-matrix $S$.
In that way a score value indicating how well the model fits the data is given separately
for each tuple of observation and estimated parameter element 
$(y_i, x_i, \hat{\beta_l})$ with $i \in \{1,\dots,n\}$
and $l \in \{1,\ldots,k\}$.

@@

If both variables $Y$ and $Z$ are numeric the applied independence test corresponds to a Pearson 
correlation test. For one of the variables being categorical and the other numeric a 1-way ANOVA
is employed while for two categorical variables a $\chi^2$-test is performed. Hence, the CTree algorithm
provides a very general framework where the choice of the type of variables and the corresponding 
functions $g$ and $h$ determine the type of the applied statistical test 
(see \cite{Hothorn+Hornik+VanDeWiel:2006} for examples on how to use the different options of this toolbox).

@@

\subsection{MOB}\label{sec:mob}

While CTree offers to apply model-based discrepancy measures as one possible choice, the 
MOB algorithm \citep{Zeileis+Hothorn+Hornik:2008} was developed with the aim of embedding 
parametric models into the framework of regression trees and is hence purely intended for 
a model-based approach.

The testing strategy for selecting a split variable is based on a fitted model, i.e. on 
an estimated parameter (vector) $\hat{\beta}$ for each node of the tree, and on the 
application of the central limit theorem. 
The fitted model can be of a very basic and simple form such as 
an average value of the response vector
%a constant fit where the parameter $\beta$ is an
%average value of the response, hence 
%$\hat{\beta} = \bar{y} = \frac{1}{n}\sum_{i=1,\ldots,n}^n y_i$ for a response vector 
%$\boldsymbol{y} = (y_1,\ldots,y_n)^\top$ 
but can also be a more complex parametric model.
In \cite{Zeileis+Hothorn+Hornik:2008} GLMs and survival regression models are used to 
exemplify the algorithm. Subsequently various other types of models have been applied as well,
for example beta regression models \citep{Gruen+Kosmidis+Zeileis:2012}, 
psychometric models such as Rasch trees \citep{Strobl+Kopf+Zeileis:2015} and
mixed effects models \citep{Fokkema+Smits+Zeileis:2018}.

Regardless of the applied model type the algorithm only requires the corresponding model 
score matrix $S$ in order to test for any instability in the model parameters over each 
particular ordering induced by each of the possible partitioning variables $Z$.
If a variable $Z$ has no specific influence on the parameters the scores would always 
fluctuate randomly around their mean 0. Therefore, these deviations are investigated by 
measuring them using an empirical fluctuation process. Applying a scalar function which 
captures the fluctuation in this process yields the test statistic of this so called 
generalized M-fluctuation test \citep{Zeileis+Hornik:2007}. 
To obtain the distribution of the resulting test statistic asymptotic theory, in particular 
the central limit theorem is applied. In this parametric unconditional framework $p$-values 
can be calculated and the covariate with the smallest $p$-value is selected as split variable.

For a categorical variable all possible splits are evaluated to find the best one while 
for $Z$ being a numeric variable a maximally selected split point search is employed. Hence, 
the number of possible breakpoints is not fixed unlike in the CTree algorithm.

\subsection{GUIDE}\label{sec:guide}
Similar to the MOB algorithm, parametric models are fit to the nodes of a tree in 
the GUIDE algorithm \citep{Loh:2002}. Again, simple models such as 
a constant fit where the parameter $\beta$ is the mean of the response, hence 
$\hat{\beta} = \bar{y} = \frac{1}{n}\sum_{i=1,\ldots,n}^n y_i$ for a response vector 
$\boldsymbol{y} = (y_1,\ldots,y_n)^\top$, can be employed as well as more complex models.
For the statistical test applied to select a split variable some type of model-based
residuals is required. In \cite{Loh:2002} linear models are used as an example where the
residuals are defined by
$$
r(Y,X,\hat{\beta}) = Y - \hat{\beta_0} - \hat{\beta_1} \cdot X.
$$
In related literature other models together with an appropriate choice of residuals
have been applied such as 
regression trees for longitudinal and multiresponse data \citep{Loh+Zheng:2013}, %(http://www.stat.rice.edu/~jrojo/4th-Lehmann/slides/Loh.pdf)
Poisson regression trees \citep{Chaudhuri+Lo+Loh:1995}, %(adjusted Anscombe residual (Pierce and Schafer 1986))
and quantile regression models \citep{Chaudhuri+Loh:2002} among many others.

Before applying a statistical test two additional steps are performed:
(1)~The residuals are binarized, i.e. depending on whether the sign of a residual is 
positive or not it is replaced by $1$ or $2$.  % FIX ME: +1 and -1 in the implementation? 
(2)~For each possible split variable the observed values are categorized into four groups 
separated by the corresponding 1st, 2nd and 3rd quartile unless the split variable is already
categorical.
Then, a $\chi^2$-test of independence is performed for the binarized residuals and
each categorized/categorical split variable.
After choosing the split variable showing the highest dependency by yielding the lowest $p$-value,
the split point minimizing the discrepancy measure is selected. The possible number of breakpoints
is determined by the categorization of the possible split variables and is in that way fixed.

\newpage
\subsection{Pruning}
\label{sec:pruning}
In early developed methods of unbiased recursive partitioning
such as FACT \citep{Loh+Vanichsetakul:1988}, QUEST \citep{Loh+Shih:1997}, 
CRUISE \citep{Kim+Loh:2001} or GUIDE \citep{Loh:2002}
the $p$-values of applied significance tests are used purely for the selection of a split variable 
based on a unified scale. All other steps are performed in the same way as in classic tree 
algorithms, such as finding the best split point by optimizing an objective function.

In subsequent approaches as in \cite{Hothorn+Hornik+Zeileis:2006}
it is shown that these $p$-values can also be used as a basis for a
pruning strategy namely prepruning: Only if the $p$-value corresponding to the selected split
variable (hence the smallest $p$-value) is smaller than a given significance level a split is
performed while otherwise no further splits are made in the current node.
In that way the information provided by the significance test is exploited in a more profound way.
However, prepruning requires the applied test to perform reasonably well as a significance test, 
i.e. it has to have power but also avoid overfitting.

For postpruning no significance level is included leading to very large trees which are then 
pruned afterwards, for example by employing cost-complexity pruning as in the 
CART algorithm \citep{Breiman+Friedman+Stone:1984}. 
Thus, for postpruning the test only needs to be able to select the best split variable but does 
not have to be effective as a significance test.

Originally the GUIDE algorithm was introduced applying postpruning while CTree and MOB
were both presented employing prepruning (see Table~\ref{tab:combinations} for a more detailed
overview). Alternatively, for model-based partitioning a pruning strategy can also use model-based 
information criteria such as AIC or BIC \citep[see for example][]{Su+Wang+Fan:2004}.
However, all testing strategies can be combined with each of these pruning strategies.
If a test performs well the choice of a pruning strategy has hardly any impact on the results, 
but as prepruning only requires already provided information it is usually less computationally
expensive than the additional calculations needed for postpruning.
However, if the test performs poorly in terms of power and overfitting, prepruning suffers from 
this deficiency while postpruning can compensate it at least partly and improve the results.
For this reason we are focusing on the testing strategies applied in tree algorithms and 
investigate how effective they are as significance tests.

%\bigskip
%\textit{Note:}
%While GUIDE originally constructs large trees and applies postpruning with the cross-validation 
%method of CART, CTree and MOB employ prepruning by setting a significance level for the 
%independence tests such that no split is performed if the lowest $p$-value is not smaller than 
%the significance level. As explained in Section~\ref{sec:evaluation}, within the simulation study 
%in Section~\ref{sec:simulation} all three algorithms are applied using prepruning in order to put 
%the focus on the power of the algorithms in terms of selecting the best split variable. 
%However, the impact of the choice of pruning strategy is further discussed in Section~\ref{sec:discussion}.

\section{Unifying Framework}\label{sec:unify}
Each of the algorithms CTree, GUIDE, and MOB can be characterized by its combination
of the type of fitted models, the type of applied tests and the employed pruning strategy
as listed in Table~\ref{tab:combinations}. These settings represent the algorithms in their
original form. However, in principle all other combinations of the listed features are possible 
as well. As explained in Section~\ref{sec:pruning} the focus of this paper is on the testing 
strategies (second column of Table~\ref{tab:combinations}). They are here discussed for the case 
of parametric models together with prepruning as the properties of this pruning strategy
are already explained/determined by the testing strategy. Subsequently combinations
with postpruning are evaluated as well to show the impact of the choice of pruning strategy.
From here on the names CTree, MOB, and GUIDE are used to refer to the algorithm-specific
testing strategies rather than the complete algorithms.

\begin{table}[t]
\centering
\caption[Table caption text]{Combinations of fitted model type, test type and pruning strategy
for the algorithms CTree, MOB, and GUIDE.}
\label{tab:combinations}
\begin{tabular}{ l c c c }
\hline 
      & Fit             & Test                     & Pruning \\ 
\hline
CTree & non-parametric  & conditional inference    & prepruning\\
MOB   & parametric      & score-based fluctuation  & prepruning (or AIC/BIC-based pruning)\\
GUIDE & parametric      & residual-based $\chi^2$  & postpruning (cost-complexity pruning)\\
\hline
\end{tabular}
\end{table}

\subsection{Unification of Testing Strategies}

Even though CTree, MOB, and GUIDE differ in the specific tests they apply (see second column of
Table~\ref{tab:combinations}) their approaches on how to select a split variable follow the same
basic structure. Thus they can be presented together in a unifying framework for testing strategies
in model-based unbiased recursive partitioning. 
In this framework each specific testing strategy can be obtained by combining the provided 
building blocks accordingly. 
This general toolbox covers all features of the here discussed testing strategies CTree, MOB, and
GUIDE. Furthermore, it allows for all possible combinations of these features and in that way also
for variations of the commonly known testing strategies. Moreover, by exchanging only one building
block its impact can be investigated in detail.
These building blocks are explained in Section~\ref{sec:buildingblocks} and their setting
for CTree, MOB, and GUIDE are listed in Table~\ref{tab:buildingblocks} which gives a more 
detailed overview of the differences in the second column of Table~\ref{tab:combinations}.

\subsection{Building Blocks of Testing Strategies}\label{sec:buildingblocks}
The three following building blocks all refer to the way how the given data is prepared/transformed
before employing a statistical test to select a split variable.
\begin{itemize}
\item  Transformations of Y:\\
A model-based transformation of $Y$ is applied to obtain a discrepancy measure giving 
information about the goodness of fit of the model. This can be residuals $r(Y,X,\hat{\beta})$
as in the GUIDE algorithm or scores $s(Y,X,\hat{\beta})$ as in the MOB algorithm or, 
more generally, a function $h(Y,X)$ as in the CTree algorithm including both of the foregoing 
cases.
By including all available scores rather than just residuals the goodness of fit of the model is 
investigated regarding each parameter instead of just the mean/location parameter. 
Therefore, this facilitates the detection of changes in parameters other than the mean/location
parameter.

\item Binarization of residuals/scores:\\
Rather than applying the independence test directly on the residuals/scores a
binarized version of them can be used as well. In particular, the values are binarized
at $0$ such that only their signs are considered.

@@

\textit{Note:}
Apart from these three building blocks, the algorithms differ in their approximation of the 
underlying null distribution of the test statistic. While the permutation tests in the CTree 
algorithm work in a conditional framework the MOB and GUIDE algorithms both apply an unconditional 
approach.

\begin{table}[t]
\centering
\caption[Table caption text]{Testing strategies of CTree, MOB, and GUIDE with the corresponding
setting of the building blocks in the unifying framework and the approximation type of the
underlying null distribution.}
\label{tab:buildingblocks}
\begin{tabular}{ l c c c c }
\hline 
& Scores            & Binarization  & Categorization  & Null distribution \\ 
\hline
CTree & full model scores & --          & --            & conditional\\
MOB   & full model scores & --          & --            & unconditional\\
GUIDE & residuals         & \checkmark  & \checkmark    & unconditional\\
\hline
\end{tabular}
\end{table}

%\newpage
